Estimating Access to Rabies Post-Exposure Prophylaxis in Madagascar

### 2019-03-14 accessPEP.Rmd
with fixed sensitivity estimates

In 2018, Gavi added PEP to their vaccine investment portfolio, and soon countries will be able to apply for funding to expand access to PEP. Expanding access strategically by identifying places with limited access...

In Madagascar, Institut Pasteur de Madagascar (IPM) provides PEP at no-cost to patients at 31 clinics across the country. PEP is not available at any other public clinics or through the private sector. In addition, there is limited control in dog populations and the disease is endemic throughout the country. Due to the spatially restricted nature of PEP, geographic access is likely to be a major driver of disease burden within the country. To get spatial estimates of disease burden in this context, we assume that bite incidence reflects access to PEP rather than than differences in rabies incidence, using travel times to clinics as a predictor of reported bites. Then using a range of rabies incidence given endemic transmission with no mass dog vaccination (*GAVI paper*), we generate sub-national estimates of rabies burden in an adapted decision tree framework. Finally, using this same model pipeline, we explore the impacts of geographically expanding access to PEP in Madagascar on reducing the burden of human rabies deaths.

@@

@@

For most districts, the majority of bites were reported to the closest clinic as estimated by our travel time metric (Fig 1A, Fig S1). Therefore, we defined the catchment area for each clinic as all districts for which the clinic was the closest ARMC. We excluded any catchment areas for which the clinic did not submit any forms (n = `r length(no_data)`, districts in grey in Fig 1A).

As even for clinics which submitted data, there was substantial undersubmission of forms, we estimated clinic level reporting as the proportion of days on which forms were submitted, excluding any periods for which there were no forms submitted for 15 consecutive days (Figure S2). Our estimate of reporting did vary based on our assumption of the threshold number of consecutive days (Figure S3), so we did look at the sensitivity of model parameter estimates to changing this threshold. To estimate the average annual bites reported for each district, we further excluded data from any years for which there was less than 25% reporting at the clinic level.

Previous work in the Moramanga District showed that low risk contacts with probable or confirmed rabies cases make up approximately 20% of patients reporting to ARMC (Rajeev et al. 2018). Generally, contacts present as clustered cases, so we excluded patients reporting on any dates that had greater than 3 standard deviations above the mean number of patients reporting per day (Figure 3A). We validated this method using the Moramanga ARMC data for which we had details on the type of exposure, and found that setting the threshold to 3 standard deviations (SDs) resulted in approximately 50% of known contacts excluded, with only 2% of non-contacts excluded (Figure S4). For the national data for which a subset of patient forms were explicitly noted to be contacts, we found that our exclusion criteria of 3 SDs identified *r round((ctar_check %$% length(exclude_bycomment[exclude_bycomment == 1 & exclude_bydate == 1]))/length(ctar_check %$% exclude_bycomment[exclude_bycomment == 1])**100, 2)"* of known contacts. We further excluded these known contacts which were not identified based on the daily distribution of patients, resulting in the exclusion of approximately *r round((ctar_check %$% length(exclude_bycomment[exclude_bycomment == 1 | exclude_bydate == 1]))/nrow(ctar_check)**100, 2)*% of patient records from the national data. We also compared this method to assuming that 20% of total bites at the district level were contacts.

@@

After excluding contacts and correcting for undersubmission of forms, our final dataset consisted of estimates of average bite incidence for `r nrow(exps_dist)` districts from *r length(unique(exps_dist$dist_catch))* catchments (Figure 4).

@@

### Model of reported bites as a function of travel time
We modeled bites as a function of travel times as follows:
$$ 
  \mu = exp(\beta_{t}T + \beta_0)P 
$$
where $mu$ is the expected number of bites given travel times and population for a given location. In this way, we can model the impact of travel time on incidence as 
$exp(\beta_{t}T + \beta_0)$ gives an estimate of incidence and multiplying by the population size gives the expected number of bites, so that we can fit this model to the number of bites in each location.

We then estimate the likelihood of observing the bites at the district level where bites are a poisson distribution around the mean $\mu_{d}$, given $\beta_t$, the effect of travel times of reported bites and $\beta_0$, the model intercept.

We fit this to our available data in three ways:
1) the commune data available at the moramanga district, where we fit to $mu_j$
2) The district data available nationally
3) To compare estimates from 1 directly to the national data, the district data but where $mu_d$ is the sum of the travel time effect at the commune level, i.e.:

$$ 
  \mu_{d} = \sum \limits_{j=1}^jexp(\beta_{t}T_j + \beta_0)P_j 
$$

 As travel times are correlated with human population size (Figure Z), we also compared how well bites were predicted by human population size alone for these different models (replacing $Tj$ with log-transformed population in $j$ to avoid squared terms), as a test of whether the observed patterns could be predicted by bite incidence scaling with population size alone. We also looked at how well distance from the closest ARMC (rather than travel times) predicted bites as an alternative proxy for access.

@@

TO DO: Instead of reporting calculate as incidence of deaths here!!
We calculated the expected reporting of rabid exposures ($\overline\rho$) given bite incidence as predicted by our model($\mu$) as:
$$
  \overline\rho = \frac{\mu \times p_{rabid}}{R}
$$
or the fraction of incidence due to rabid exposures ($\mu \times p_{rabid}$) divided by the total rabies exposure incidence ($r$) for a range of estimated rabies incidence and $p_{rabid}$. We look at the range of $p_{rabid}$ reported in Rajeev et al. 2018 for data from the Moramanga District (0.2 - 0.6). where the proportion of reported bites that are rabies exposures ($p_{rabid}$) are defined as:
$$
           p_{rabid}=
            \begin{cases}
            x, & \text{if}\ \frac{R \times \rho_{max}}{B} > x \\
            \frac{R \times \rho_{max}}{B}, & \text{otherwise}
            \end{cases}
$$
<br>

such that rabid reported bites (i.e. $p_{rabid} \times B_i$) cannot exceed the expected number of human exposures given maximum reporting (i.e. $R_i \times \rho_{max}$). $\rho_{max}$ is taken from the Moramanga ARMC data for Moramanga Ville, the commune with the ARMC (i.e. the area with the minimum travel time in the district, on average *xx* minutes).

@@

To generate $R$, the rabies incidence in dogs in the absence of any vaccination, $r$, is  multiplied by the estimated dog population in the commune ($D_i$) and the exposure rate per rabid dogs ($p_{exp}$ = 0.39 persons exposed per rabid dog)(Hampson et al. 2018). We use a human:dog ratio (HDR) of 5 to generate our maximum expected incidence and an HDR of 25 for our minimum expected incidence. As there is little data on dog populations in Madagascar, this range of HDRs encompasses a wide range observed across Africa (cite!).

To estimate the average number of deaths for each commune, we extended the above framework into a stochastic framework as follows):
$$
deaths_i = (R_i - p_{rabid_i}B_i) \times p_{death}
$$
where $R_i$ is drawn from a uniform distribution between the minimum and maximum expected number of human exposures for each location and $B_i$, the number of reported bites, is drawn from a poisson distribution with the mean predicted number of bites from the travel time model. We draw $p_rabid_i$ from a uniform distribution between 0.2 - 0.6, while constraining it as per Equation N. We assume that all rabies exposed patients who report to an ARMC receive and complete PEP, and PEP is completely effective at preventing death due to rabies. The probability of death in the absence of PEP is taken from cite GAVI/Joel paper.  
<br>

@@

### Sensitivity analyses for burden estimates 
We estimate burden deterministically across a range of parameter values to test the effects of our model assumptions on estimates of rabies burden. Specifically, we fix rabies incidence at the minimum and maximum of our estimated range, look at the range of values of p_rabid between 0.2 - 0.6 (as per Moramanga), and the range of values of rho_max (0.85 - 0.99) to get at maximum and minimum estimates of burden. We also examine the impact of systematic variation in rabies incidence with human population (a potential proxy for changes in the dog population/HDR) by looking at how estimates of burden changes if rabies incidence were to scale positively or negatively with human population.

@@

ggplot(preds_fixed, aes(x = covar, y = predicted, color = scale)) + 
  geom_line() + 
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = scale), color = NA, alpha = 0.5) +
  scale_color_manual(values = c("navy", "blue", "purple")) +
  scale_fill_manual(values = c("blue", "lightblue", "pink")) +
  xlab("Travel time (hrs)") +
  ylab("Reported bites per \n 100,000 persons")

ggplot(preds_data, aes(x = observed, y = predicted, color = scale)) + 
  geom_point(alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "grey") +
  expand_limits(y = c(0, 4000)) +
  scale_color_manual(values = c("navy", "blue", "purple"))

@@

### Estimation of burden and reporting

Generally, estimated reporting of rabies exposures decayed with travel times given model predicted bite incidence and a range of rabies incidence and $p_{rabid}$ (Figure 6). Given our model assumptions, reporting was estimated at the maximum of 0.98 for travel times under 1 hour given the maximum estimated rabies exposure incidence and the minimum estimate of $p_{rabid}$ (the lower range of reporting probabilities), and travel times under 5.5 hours given the minimum estimated rabies exposure incidence and the maximum estimate of $p_{rabid}$ (the upper range of reporting probabilities).

```{r burden results, fig.cap = "Figure 7", cache = TRUE}
# Incremental analysis ------------------------------------------------------------------------
## District travel times
dist_ttimes <- read.csv("output/district_temp_scenario_20190212_181056.csv", row.names = 1)
comm_ttimes <- read.csv("output/commune_temp_scenario_20190212_182810.csv", row.names = 1)
B_ttimes_mada <- bite_mods$beta[bite_mods$model == "Mada flatPop ttimes_commune"]
B_0_mada <- bite_mods$intercept[bite_mods$model == "Mada flatPop ttimes_commune"]

@@

@@

## Group commune results to district level
colnames(results_comm_med) <- gsub("results.", "", colnames(results_comm_med))
results_comm_med %>%
  left_join(select(comm_data, mdg_dis_co, mdg_cm_), 
            by = c("names" = "mdg_cm_")) %>%
  group_by(scenario, mdg_dis_co) %>%
  summarize(deaths_mean = sum(deaths_mean, na.rm = TRUE), 
            deaths_upperCI = sum(deaths_upperCI, na.rm = TRUE),
            deaths_lowerCI = sum(deaths_lowerCI, na.rm = TRUE)) %>%
  rename(names = mdg_dis_co) -> results_comm_summarized
results_comm_summarized$model <- "commune"
colnames(results_dist_med) <- gsub("results.", "", colnames(results_dist_med))
results_dist_med %>%
  mutate(model = "district") %>%
  select(scenario, names, deaths_mean, deaths_lowerCI, deaths_upperCI, 
         model) -> results_dist_med
results_all <- bind_rows(results_dist_med, results_comm_summarized)
results_all %>%
  left_join(select(mada_district@data, district, mdg_dis_co, ttimes_weighted),
            by = c("names" = "mdg_dis_co")) -> results_all

## Plot national level burden
results_all %>% 
  group_by(scenario, model) %>% 
  summarize(deaths_mean = sum(deaths_mean, na.rm = TRUE), 
            deaths_upperCI = sum(deaths_upperCI, na.rm = TRUE),
            deaths_lowerCI = sum(deaths_lowerCI, na.rm = TRUE)) -> results_natl

## Plotting national level
ggplot(data = results_natl, aes (x = scenario, y = deaths_mean, color = model)) +
  geom_line() +
  geom_ribbon(aes(ymin = deaths_upperCI, ymax = deaths_lowerCI, 
                  color = NULL, fill = model),
              alpha = 0.5) +
  scale_color_manual(values = c("district" = "blue", 
                                "commune" = "purple"), 
                     guide = "none") +
  scale_fill_manual(values = c("district" = "blue", 
                                "commune" = "purple")) +
  guides(color = guide_legend(), 
         fill = guide_legend(override.aes = list(linetype = 1)))

@@

@@

When we estimate burden of deaths stochastically within this range of incidence and given our high and low estimates of proportion of reported bites that are rabid, we see that burden of deaths also decreases with travel times (Figure 7, results presented aggregated at the district level). Overall, we estimate average annual deaths between .... **Also estimate deaths averted here!**.

When we compare our burden estimates at the district vs. the commune level (summed to district), we see that while overall, the estimates are very similar (Fig NA), at low travel times, calculating burden at the district level results in an assumption of maximum reporting for the whole district, which assumes very low burden.

>>>>

In Madagascar, Institut Pasteur de Madagascar (IPM) provides PEP at no-cost to patients at 31 clinics, or anti-rabies medical centers (ARMC) across the country. PEP is not available at any other public clinics or through the private sector. In addition, there is limited control in dog populations and the disease is endemic throughout the country. Due to the spatially restricted nature of PEP, geographic access is likely to be a major driver of disease burden within the country. To get spatial estimates of disease burden in this context, we use data on reported bite incidence at the sub-national level (district) and assume that bite incidence reflects access to PEP rather than than differences in rabies incidence. We fit the data to travel times to the closest ARMC, a metric of geographic access. Then using a range of rabies incidence given endemic transmission with no mass dog vaccination (*GAVI paper*), we generate sub-national estimates of rabies burden in an adapted decision tree framework.

 In 2018, Gavi added PEP to their vaccine investment portfolio, and by 2020, Gavi-eligible countries will be able to apply for funding to expand access to PEP, which has the potential to drastically reduce burden of rabies. We explore the impact of this potential intervention using this same model pipeline, we explore the impacts of geographically expanding access to PEP in Madagascar on reducing the burden of human rabies deaths.

@@

## Getting other data
rep_data <- get.reporting(ctar_data, ctar_metadata, start_date = ymd("2014-01-01"), 
                          end_date = ymd("2017-12-31"), 
                          threshold = 10)
ctar_contacts <- get.contacts(ctar_data, times_sd = 3)
ctar_contacts <- ctar_contacts[["ctar_data"]]
```

@@

For most districts, the majority of bites were reported to the closest clinic as estimated by our travel time metric (Fig 1A, Fig S1). Therefore, we defined the catchment area for each clinic as all districts for which the clinic was the closest ARMC. We excluded any catchment areas for which the clinic did not submit any forms (catchments = `r length(no_data)`, districts = `r nrow(exps_dist) - nrow(mada_district)`, in grey in Fig 1A).

As even for clinics which submitted data, there was substantial undersubmission of forms, we estimated clinic level reporting as the proportion of days on which forms were submitted, excluding any periods for which there were no forms submitted for 15 consecutive days (Figure S2). Our estimate of reporting did vary based on our assumption of the threshold number of consecutive days (Figure S3), so in subsequent analyses we looked at the sensitivity of model parameter estimates to changing this threshold. To estimate the average annual bites reported for each district, we further excluded data from any years for which there was less than 25% reporting at the clinic level.

Previous work in the Moramanga District showed that low risk contacts, i.e. touching or feeding a suspected animal or human rabies case, make up approximately 20% of patients reporting to ARMC (Rajeev et al. 2018). These low risk exposures do not contribute to burden and therefore we attempted to exclude them from our analysis. Generally, contacts present as clustered cases, so we excluded patients reporting on any dates that had greater than 3 standard deviations above the mean number of patients reporting per day (Figure 3A). We validated this method using the Moramanga ARMC data for which we had details on the type of exposure, and found that setting the threshold to 3 standard deviations (SDs) resulted in approximately 50% of known contacts excluded, with only 2% of non-contacts excluded (Figure S4). For the national data for which a subset of patient forms were explicitly noted to be contacts, we found that our exclusion criteria of 3 SDs identified `r round((ctar_contacts %$% length(exclude_bycomment[exclude_bycomment == 1 & exclude_bydate == 1]))/length(ctar_contacts %$% exclude_bycomment[exclude_bycomment == 1])*100, 2)` of known contacts. We further excluded these known contacts which were not identified based on the daily distribution of patients, resulting in the exclusion of approximately `r round((ctar_contacts %$% length(exclude_bycomment[exclude_bycomment == 1 | exclude_bydate == 1]))/nrow(ctar_contacts)*100, 2)` % of patient records from the national data.

@@

After excluding contacts and correcting for undersubmission of forms, our final dataset consisted of estimates of average bite incidence for `r nrow(exps_dist)` districts from `r length(unique(exps_dist$dist_catch))` catchments (Figure 4).

@@

### Model of reported bites as a function of travel time
We modeled incidence of bites as a function of travel times in hours as follows:
$$ 
  \mu = exp(\beta_{t}T + \beta_0)P 
$$
where $mu$ is the expected number of bites given travel times and population for a given location. We then estimate the likelihood of observing the bites at the district level where bites are a poisson distribution around the mean $\mu_{d}$, given $\beta_t$, the effect of travel times of reported bites and $\beta_0$, the model intercept.

We fit this to our available data in three ways:
1) Commune level bite incidence from the Moramanga District
2) District level bite incidence as calculated above from the national database
3) District level bite incidence, but with commune level travel times ($T_j$) so that:

$$ 
  \mu = \sum \limits_{j=1}^jexp(\beta_{t}T_j + \beta_0)P_j 
$$

As travel times are correlated with human population size (Figure S4), we also compared how well bites were predicted by human population size. We used a generalized linear model with a possion response-now predicting the number of bites rather than bite incidence` with either population size alone or in combination with travel times, so that incidence could scale with population size. We also looked at how well distance from the closest ARMC (rather than travel times) predicted bites as an alternative proxy for access, replacing travel times in hours ($T$) with distance from the nearest ARMC in km ($D$). Distance to the nearest ARMC was taken as the minimum distance from the centroid of the commune or district to any ARMC.

@@

We calculated the expected reporting of rabid exposures ($\overline\rho$) given bite incidence as predicted by our model($\mu$) as:
$$
  \overline\rho = \frac{\mu \times p_{rabid}}{R}
$$
or the fraction of incidence due to rabid exposures ($\mu \times p_{rabid}$) divided by the total rabies exposure incidence ($r$) for a range of estimated rabies incidence and $p_{rabid}$. We look at the range of $p_{rabid}$ reported in Rajeev et al. 2018 for data from the Moramanga District (0.2 - 0.6). where the proportion of reported bites that are rabies exposures ($p_{rabid}$) are defined as:
$$
           p_{rabid}=
            \begin{cases}
            x, & \text{if}\ \frac{R \times \rho_{max}}{B} > x \\
            \frac{R \times \rho_{max}}{B}, & \text{otherwise}
            \end{cases}
$$
<br>

such that rabid reported bites (i.e. $p_{rabid} \times B_i$) cannot exceed the expected number of human exposures given maximum reporting (i.e. $R_i \times \rho_{max}$). $\rho_{max}$ is taken from the Moramanga ARMC data for Moramanga Ville, the commune with the ARMC (i.e. the area with the minimum travel time in the district, on average `r round(exps_mora$ttimes_weighted[exps_mora$commune == "Moramanga"], 2)` minutes).

@@

To generate $R$, the rabies incidence in dogs in the absence of any vaccination, $r$, is  multiplied by the estimated dog population in the commune ($D_i$) and the exposure rate per rabid dogs ($p_{exp}$ = 0.39 persons exposed per rabid dog)(Hampson et al. 2018). We use a human:dog ratio (HDR) of 5 to generate our maximum expected incidence and an HDR of 25 for our minimum expected incidence, which gives a range of approximately `r 1e5/25*0.01*0.39` - `r 1e5/5*0.01*0.39` exposures per 100,000 persons. As there is little data on dog populations in Madagascar, this range encompasses both a wide range of observed HDRs and exposure rates across Africa (cite!).

To estimate the average number of deaths for each commune, we extended the above framework into a stochastic framework as follows:
$$
deaths_i = (R_i - p_{rabid_i}B_i) \times p_{death}
$$
where $R_i$ is drawn from a uniform distribution between the minimum and maximum expected number of human exposures for each location and $B_i$, the number of reported bites, is drawn from a poisson distribution with the mean predicted number of bites from the travel time model. We draw $p_rabid_i$ from a uniform distribution between 0.2 - 0.6, while constraining it as per Equation N. We assume that all rabies exposed patients who report to an ARMC receive and complete PEP, and PEP is completely effective at preventing death due to rabies. The probability of death in the absence of PEP is taken from cite GAVI/Joel paper.  
<br>

@@

We also look at how this may impact vaccine demand. Specifically, we fit the observed daily distribution of patients receiving PEP at each clinic to a negative binomial distribution, with both the dispersion parameter and mean as a function of the annual average number of reported bites. We use this to simulate clinic throughput and vials used as clinics are added, assuming routine vial sharing practices used currently in Madagascar (i.e. each vial is split between two patients, and vials are discarded at the end of each day). We project the potential increase in vial demand and change in the ratio of number of vials per death averted.

### Sensitivity analyses for burden estimates 
We estimate burden deterministically across a range of parameter values to test the effects of our model assumptions on estimates of rabies burden. Specifically, we fix rabies incidence at the minimum and maximum of our estimated range, look at the range of values of p_rabid between 0.2 - 0.6 (as per Moramanga), and the range of values of rho_max (0.85 - 0.99) to get at maximum and minimum estimates of burden. We also examine the impact of systematic variation in rabies incidence with human population (a potential proxy for changes in the dog population/HDR) by looking at how estimates of burden change if rabies incidence were to scale positively or negatively with human population.

@@

library(patchwork)
a <- ggplot(preds_fixed, aes(x = covar, y = predicted, color = scale)) + 
  geom_line() + 
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = scale), color = NA, alpha = 0.5) +
  scale_color_manual(values = c("navy", "blue", "purple")) +
  scale_fill_manual(values = c("blue", "lightblue", "pink"), name = "Model scale") +
  xlab("Travel time (hrs)") +
  ylab("Reported bites per \n 100,000 persons") +
  labs(tag = "A")

b <- ggplot(preds_data, aes(x = observed, y = predicted, color = scale)) + 
  geom_point(alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "grey") +
  expand_limits(y = c(0, 4000)) +
  scale_color_manual(values = c("navy", "blue", "purple"), name = "Model scale") +
  xlab("Mean annual bites observed") +
  ylab("Mean annual bites predicted") +
  labs(tag = "B")
a / b

kable(bite_mods)
```

@@

The best fitting model did not include population as a predictor for the district data...write up from supplement!

### Estimation of burden and reporting 
```{r}
B_ttimes_mada <- bite_mods$beta[bite_mods$model == "Mada flatPop ttimes_commune"]
B_0_mada <- bite_mods$intercept[bite_mods$model == "Mada flatPop ttimes_commune"]

@@

ttimes <- seq(0, 1200, by = 60)/60
p_rabid_vals <- c(0.2, 0.6)
hdr_vals <- c(5, 25)
beta_vals <- list(District = B_ttimes_dist, Commune = B_ttimes_mada)
int_vals <- list(District = B_0_dist, Commune = B_0_mada)

rep_df <- 
  foreach(i = 1:length(p_rabid_vals), .combine = rbind) %:% 
  foreach(k = 1:length(hdr_vals), .combine = rbind) %:%
  foreach(j = 1:length(beta_vals), .combine = rbind) %do% {
    rep <- as.data.frame(list(p_rabid = p_rabid_vals[i], ttimes = ttimes, hdr = hdr_vals[k], 
                                         reporting = get.burden.fixed(pop = 1e5, ttimes = ttimes,
                                                          B_ttimes = beta_vals[[j]], B_0 = int_vals[[j]],
                                            hdr = hdr_vals[k], p_rabid = p_rabid_vals[i],
                                            type = "reporting", scale = FALSE), scale = names(beta_vals)[j]))
  }

rep_min_max <- 
  foreach(j = 1:length(beta_vals), .combine = rbind) %do% {
    rep_max <- get.burden.fixed(pop = 1e5, ttimes = ttimes, B_ttimes = beta_vals[[j]], B_0 = int_vals[[j]],
                                            hdr = 25, p_rabid = 0.6,
                                            type = "reporting", scale = FALSE)

    rep_min <- get.burden.fixed(pop = 1e5, ttimes = ttimes, B_ttimes = beta_vals[[j]], B_0 = int_vals[[j]],
                                            hdr = 5, p_rabid = 0.2,
                                            type = "reporting", scale = FALSE)
    rep <- as.data.frame(list(ttimes = ttimes, rep_max = rep_max, rep_min = rep_min, 
                              scale = names(beta_vals)[j]))
  }

ggplot(rep_df, aes(x = ttimes, y = reporting, color = interaction(p_rabid, hdr), linetype = as.factor(hdr))) +
  geom_line() +
  facet_wrap(~ scale) +
  scale_color_manual(values = c("red", "red", "blue", "blue"))

ggplot(rep_min_max, aes(x = ttimes, y = rep_min, fill = scale)) +
  geom_line(aes(y = rep_min, color = scale), linetype = 2) +
  geom_line(aes(y = rep_max, color = scale), linetype = 1) +
  geom_ribbon(aes(ymax = rep_max, ymin = rep_min), alpha = 0.5) +
  scale_fill_manual(values = c("navy", "blue")) +
  scale_color_manual(values = c("navy", "blue"))

Generally, estimated reporting of rabies exposures decayed with travel times given model predicted bite incidence and a range of rabies incidence and $p_{rabid}$ (Figure N). Given our model assumptions, reporting was estimated at the maximum of 0.98 for travel times under 1 hour given the maximum estimated rabies exposure incidence and the minimum estimate of $p_{rabid}$ (the lower range of reporting probabilities), and travel times under 5.5 hours given the minimum estimated rabies exposure incidence and the maximum estimate of $p_{rabid}$ (the upper range of reporting probabilities). We found that the commune model of bites estimated a lower range of reporting probabilities that the district model (Figure SX), so we compared estimates of these two models in subsequent analyses.

```{r burden results, fig.cap = "Figure 7", cache = TRUE}
# Incremental analysis ------------------------------------------------------------------------
## District travel times
dist_ttimes <- read.csv("output/district_temp_scenario_20190212_181056.csv", row.names = 1)
comm_ttimes <- read.csv("output/commune_temp_scenario_20190212_182810.csv", row.names = 1)

@@

## Group commune results to district level
colnames(results_comm_med) <- gsub("results.", "", colnames(results_comm_med))
results_comm_med %>%
  left_join(select(comm_data, mdg_dis_co, mdg_cm_), 
            by = c("names" = "mdg_cm_")) %>%
  group_by(scenario, mdg_dis_co) %>%
  summarize(deaths_mean = sum(deaths_mean, na.rm = TRUE), 
            deaths_upperCI = sum(deaths_upperCI, na.rm = TRUE),
            deaths_lowerCI = sum(deaths_lowerCI, na.rm = TRUE),
            averted_mean = sum(averted_mean, na.rm = TRUE), 
            averted_upperCI = sum(averted_upperCI, na.rm = TRUE),
            averted_lowerCI = sum(averted_lowerCI, na.rm = TRUE)) %>%
  rename(names = mdg_dis_co) -> results_comm_summarized
results_comm_summarized$model <- "commune"
colnames(results_dist_med) <- gsub("results.", "", colnames(results_dist_med))
results_dist_med %>%
  mutate(model = "district") %>%
  select(scenario, names, deaths_mean, deaths_lowerCI, deaths_upperCI, 
         averted_mean, averted_lowerCI, averted_upperCI,
         model) -> results_dist_med
results_all <- bind_rows(results_dist_med, results_comm_summarized)
results_all %>%
  left_join(select(mada_district@data, district, mdg_dis_co, ttimes_weighted),
            by = c("names" = "mdg_dis_co")) -> results_all

## Plot national level burden
results_all %>% 
  group_by(scenario, model) %>% 
  summarize(deaths_mean = sum(deaths_mean, na.rm = TRUE), 
            deaths_upperCI = sum(deaths_upperCI, na.rm = TRUE),
            deaths_lowerCI = sum(deaths_lowerCI, na.rm = TRUE), 
            averted_mean = sum(averted_mean, na.rm = TRUE), 
            averted_upperCI = sum(averted_upperCI, na.rm = TRUE),
            averted_lowerCI = sum(averted_lowerCI, na.rm = TRUE)) -> results_natl

## Deaths averted
ggplot(results_all_base, aes(y = reorder(district, ttimes_weighted), x = averted_mean, 
                              color = model)) +
  ggstance::geom_pointrangeh(aes(xmin = averted_lowerCI, xmax = averted_upperCI), fatten = 2) +
  ylab("") +
  xlab("Average annual deaths") +
  scale_color_manual(values = c("district" = alpha("blue", 0.5), 
                                "commune" = alpha("purple", 0.5)), 
                     name = "Scale") -> p
p + theme(axis.text.y = element_text(margin = margin(t = 10, r = 0, b = 10, l = 0), size = 7))

kable(results_all_base %>%
  group_by(model) %>%
  summarize(averted = sum(averted_mean), 
            deaths = sum(deaths_mean)) -> summary_burden)

When we estimate burden of deaths stochastically within this range of incidence and of proportion of reported bites that are rabid, we see that burden of deaths also decreases with travel times (Figure XA, results presented aggregated at the district level). Overall, we estimate average annual deaths between `r min(summary_burden$deaths)` - `r max(summary_burden$deaths)` and deaths_averted between `r min(summary_burden$averted)` - `r max(summary_burden$averted)`.

Our sensitivity analyses show, that while there is considerable uncertainty in the number of deaths estimated, the pattern of deaths increasing with travel times is robust...write up from supplement!

``` {r scenario}
## Plotting national level
ggplot(data = results_natl, aes (x = scenario, y = deaths_mean, color = model)) +
  geom_line() +
  geom_ribbon(aes(ymin = deaths_upperCI, ymax = deaths_lowerCI, 
                  color = NULL, fill = model),
              alpha = 0.5) +
  scale_color_manual(values = c("district" = "blue", 
                                "commune" = "purple"), 
                     guide = "none") +
  scale_fill_manual(values = c("district" = "blue", 
                                "commune" = "purple")) +
  guides(color = guide_legend(), 
         fill = guide_legend(override.aes = list(linetype = 1)))

@@

## Deaths averted
ggplot(data = results_natl, aes (x = scenario, y = averted_mean, color = model)) +
  geom_line() +
  geom_ribbon(aes(ymin = averted_upperCI, ymax = averted_lowerCI, 
                  color = NULL, fill = model),
              alpha = 0.5) +
  scale_color_manual(values = c("district" = "blue", "commune" = "purple"),  guide = "none") +
  scale_fill_manual(values = c("district" = "blue", "commune" = "purple")) +
  guides(color = guide_legend(), fill = guide_legend(override.aes = list(linetype = 1)))

ggplot(data = results_all, aes(x = averted_mean, y = reorder(district, ttimes_weighted), 
                               color = scenario)) +
  ggstance::geom_pointrangeh(aes(xmin = deaths_lowerCI, xmax = deaths_upperCI), alpha = 0.25,
                             fatten = 2) +
  facet_wrap(~model)

@@

We find that both expanding access to PEP reduces burden of rabies death, but this effect decays as ARMC are added, saturating after about approximately 200 clinics (or 2 ARMC per district). This effect is more pronouned for the commune level model than for the district model. Sensitivity analyses...write up from supplement!

### Estimating vaccine demand with increased access

```{r vial analysis}
# ## new ctar
rep_data <- get.reporting(ctar_data, ctar_metadata, start_date = ymd("2014-01-01"), 
                          end_date = ymd("2017-12-31"), 
                          threshold = 10)
date_mat <- rep_data[["date_mat"]]
dose_mat <- rep_data[["doses_wide"]]
colnames(date_mat) <- colnames(dose_mat)[2:(ncol(dose_mat)-1)]
doses <- melt(dose_mat, id.vars = "date_de_consultation", value.name = "patients", variable.name = "id_ctar")

date_mat %>% 
  as_tibble() %>%
  mutate(date = dose_mat$date_de_consultation) -> date_mat
exclude <- melt(date_mat, id.vars = "date", value.name = "exclude", variable.name = "id_ctar")
doses %>%
  left_join(exclude, by = c("date_de_consultation" = "date", "id_ctar" = "id_ctar")) -> doses
doses %>%
  filter(exclude == 1) %>%
  group_by(id_ctar) %>%
  mutate(sum = sum(patients, na.rm = TRUE), mean = mean(patients, na.rm = TRUE)) %>%
  filter(sum > 10) -> doses

ggplot(data = doses, aes(x = date_de_consultation, y = id_ctar)) +
  geom_tile(aes(fill = patients))

library(ggridges)
ggplot(doses, aes(x = patients, y = id_ctar)) +
  geom_density_ridges()

doses %>%
  split(.$id_ctar) %>% # from base R
  map(~ fitdistr(.$patients, "negative binomial")) -> dist_ts

coeff_df <- as.data.frame(list(mu = mapply(function(x) coef(x)["mu"], dist_ts),
                                k = mapply(function(x) coef(x)["size"], dist_ts),
                                mu_sd = mapply(function(x) x$sd["mu"], dist_ts),
                                k_sd = mapply(function(x) x$sd["size"], dist_ts)))
coeff_df$id_ctar <- names(dist_ts)
ggplot(data = coeff_df, aes(x = mu, y = k)) + geom_point()

exps_dist %>%
  left_join(select(ctar_metadata, CTAR, id_ctar), by = c("dist_catch" = "CTAR")) %>%
  mutate(id_ctar = as.character(id_ctar)) %>%
  group_by(id_ctar) %>%
  summarize(bites = sum(bites)) %>%
  left_join(coeff_df, by = c("id_ctar" = "id_ctar")) -> coeff_df
ggplot(data = coeff_df, aes(x = bites, y = mu)) + geom_point()
ggplot(data = coeff_df, aes(x = bites, y = k)) + geom_point()

exps_dist %>%
  left_join(select(ctar_metadata, CTAR, id_ctar), by = c("dist_catch" = "CTAR")) %>%
  mutate(id_ctar = as.character(id_ctar)) %>%
  group_by(id_ctar) %>%
  summarize(bites = sum(bites)) %>%
  left_join(doses, by = c("id_ctar" = "id_ctar")) -> doses

get.throughput <- function(patient_ts, bites, B_k, B_0_k, B_mu, B_0_mu) {
  mu <- exp(B_mu*bites/100 + B_0_mu) 
  k <- exp(B_k*bites/100 + B_0_k)
  return(-sum(dnbinom(patient_ts, size = k, mu = mu, log = TRUE)))
}

fit <- mle2(get.throughput, start = list(B_0_mu = 0.1, B_k = 0.1, B_mu = 0.1, B_0_k = 0.1),
             data = list(patient_ts = doses$patients, bites = doses$bites))

## Get the observed + predicted distribution for each set of coeffs
check <- foreach(i = iter(coeff_df, by = "row"), .combine = rbind) %do% {
  check <- rnbinom(n = 365, mu = exp(0.029*i$bites/100 + -0.307),
                  size = exp(0.022*i$bites/100 + -0.39))
  as.data.frame(list(patients = check, type = "predicted", id_ctar = i$id_ctar))
}

doses %>%
  select(id_ctar, patients) %>%
  mutate(type = "observed") %>%
  bind_rows(check) -> compare_dists

compare_dists %>%
  group_by(type, id_ctar, patients) %>%
  summarize(count = n()) %>%
  group_by(type, id_ctar) %>% # next get # of days total (should always be 365 for predicted)
  mutate(total = sum(count), prop = count/total) -> compare_dists
compare_dists %>%
  mutate(prop_plot = ifelse(type == "predicted", -prop, prop)) -> compare_dists

ggplot(compare_dists, aes(x = patients, y = prop_plot, fill = type)) +
  geom_col(color = NA, alpha = 0.5, position = "identity") +
  scale_fill_manual(values = c("red", "blue")) +
  facet_wrap(~id_ctar, scales = "free") +
  xlab("# of Patients") +
  ylab("Proportion of days")

rep_bites = seq(0, 15000, by = 100)
predicts <- as.data.frame(list(rep_bites = rep_bites,
                               predicted_mu = exp(0.029*rep_bites/100 + -0.307),
                               predicted_k = exp(0.022*rep_bites/100 + -0.39)))
ggplot(predicts, aes(x = rep_bites, y = predicted_mu)) +
  geom_line()
ggplot(predicts, aes(x = rep_bites, y = predicted_k)) +
  geom_line()

We fit the distributions of patients reporting per day to the estimated average annual bites reported to each clinic. We found that as reported bites increase, the average number of patients reporting per day increases linearly, and that the heterogeneity in this increases with lower reported bites to a clinic.

We use this to estimate how vaccine demand will change as access to PEP is expanded...
